Efficient Simplex Parameterizations for Hamiltonian Monte Carlo

### 2022-11-11 paper.tex
more updates

In this paper, we focus on transforms on a unit $N$-simplex. Simplexes are useful for representations of multinomial probabilities - points on the standard $N$-simplex in $R^{N+1}$ are the space of possible parameters (probabilities) of the categorical distribution on $n+1$ possible outcomes. Lastly, we define Aitchison Geometry and compare metric distortions of canonical transforms (or their augmented forms) in Compositional Data Analysis.
\section{Constraining transforms}

A surjective constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$ is mapped from an unconstrained space $\mathcal{Y} = \mathbb{R}^M$ onto a constrained space $\mathcal{X} \subset \mathbb{R}^N$, and $h(y)$ is a proper density defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.  It is not possible to sample over an improper density.  To this end, we stipulate the induced distribution on $y$ to be proper, given a proper density $p_X(x)$ defined for $x \in \mathcal{X}$ such that if $y \sim p_Y(\cdot)$ then $f(y) \sim p_X(\cdot)$ . In this case, density of the distribution induced on $y$ is 
\[
p_Y(y) = h(y) \, p_X(f(y))
\]

@@

 Suppose, we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$. A smooth, bijective unconstraining transform is defined as $g : \mathcal{X} \rightarrow \mathcal{Y}$, and $g^{-1}$ is the constraining transform $f : \mathcal{Y} \rightarrow \mathcal{X}$.  Using the standard change of variables, the induced density on $\mathcal{Y}$ is adjusted by a constant factor

@@

For a constraining transform $f: R^\mathcal{M} \rightarrow R^\mathcal{N}$, if $\mathcal{M} >  \mathcal{N}$, additional constraints are needed to ensure the correct distributions on $\mathcal{X}$. Bijective change-of-variables method cannot be applied here, instead an expanded parameter space $\mathcal{Z} = \mathcal{X} \times \mathcal{R}$ is constructed with parameters $z = (x, r)$. Here, $r \in \mathcal{R}$ is introduced such that $\mathcal{Y}$ and $\mathcal{Z}$ have the same dimensionality. Taking prior $p_R(r | x)$,
\[
p_Y(y) = h(y) p_Z(z) = h(y) p_X(x) p_R(r | x).
\]
We define a bijective transform $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, this permits obtaining change-of-variables,
\[
h(y) = |J_{\tilde{f}}(y)|.
\]
Discarding $r$ after sampling is equivalent to marginalizing it out, and in practice, $r$ may never be directly computed. On acount of convenience,  we choose $p_R(r | x) = p_R(r)$ in the rest of the paper.

% In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transform coupled with a standard normal distribution over $\mathcal{Y}$,
% \[
%   h(y) = \textrm{normal}(y \mid 0, \textrm{I}) \, \absdet{J_f(y)},
% \]  
% where $\textrm{I}$ is the identity matrix.
% For each case, we have to ensure that $p_Y(y) = h(y) \, p_X(f(y))$ is such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.

\section{Unit simplex}

@@

Compositional data analysis deals with vectors of strictly positive quantities that adhere to a constant sum constraint. \cite{aitchison1982statistical} defined compositional data on a simplical space:
$$
\Delta^N=\left\{\mathbf{x}=\left[x_1, x_2, \ldots, x_D\right] \in \mathbb{R}^N \mid x_i>0, i=1,2, \ldots, N ; \sum_{i=1}^N x_i=\kappa\right\}
$$

@@

\subsubsection{Additive log ratio transform}
The unconstraining transform for the identified softmax is known as
the additive log ratio (ALR) transform
\cite{aitchison1982statistical}, which is a bijection
$\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
$x \in \Delta^{N-1}$ by
\[
  \textrm{alr}(x)
  = \begin{bmatrix}\displaystyle
    \log \frac{x_1}{x_N}, \cdots , \log \frac{x_{N-1}}{x_N}
  \end{bmatrix}
\]
The inverse additive log ratio transform maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{alr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,

@@

\[
\textrm{Here, } \abs{J} = \prod \exp(y)
  \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
\]

Geometrically, coordinates obtained from an Additive Log Ratio transform do not correspond to an orthogonal basis, rather it is an oblique basis with respect to the Aitchison metric. Also, the transform is not symmetrical in all the N components, rather only the first N-1. This distorts euclidean distances measured from the ALR coordinates
\subsubsection{Stick-Breaking Transform}

The stick-breaking transform is derived from the standard construction for Dirichlet processes \cite{sethuraman1994constructive}. The inverse mapping from unconstrained to constrained parameters involves taking a unit-length stick and recursively breaking pieces off until it is divided into $N + 1$ pieces, the lengths of which provide the entries in the simplex.  Following the transforms used by Stan \cite{stan2022ref}, we add an offset so that the unconstraining transform of the uniform simplex $x = \begin{bmatrix} 1/N & \cdots & 1/N \end{bmatrix}$ maps to $y = \begin{bmatrix} 0 & \cdots & 0 \end{bmatrix}$, a vector of zeros.

\paragraph{Unconstraining transform}
The unconstraining transform $f : \Delta^{N} \to  \mathbb{R}^N$ is defined so that if $f(x) = y$, then
\[
y_n
= \mathrm{logit}(z_n)
  - \mbox{log}\left(\frac{1}{N + 1 - n} \right),
\]
where 
\[ 
 z_n = \frac{x_i}{1 - \sum_{i = 1}^{n-1} x_{i}}.
\]

\paragraph{Constraining inverse transform}
The inverse transform follows the stick breaking construction, with $f^{-1} \colon \mathbb{R}^N \to \Delta^N$ defined by $f^{-1}(y) = x$, where
\[
x_n = z_n \left( 1 - \sum_{i=1}^{n-1} x_i \right), \\
\]
with
\[
z_n = \mathrm{logit}^{-1}\!\left(
y_n  + \log \left( \frac{1}{N - n} \right)
\right).
\]

\paragraph{Change-of-variables adjustment} 
The absolute determinant of the Jacobian of the inverse transform is
\[
\abs{J_{f^{-1}}} 
= \prod_{n=1}^{N-1}
   z_n \, (1 - z_n)
   \left( 1 - \sum_{i=1}^{n-1} x_{i} \right).
\]

% \subsubsection{Additive log ratio transform}
% The unconstraining transform for the identified softmax is known as
% the additive log ratio (ALR) transform
% \cite{aitchison1982statistical}, which is a bijection
% $\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
% $x \in \Delta^{N-1}$ by
% \[
%   \textrm{alr}(x)
%   = \begin{bmatrix}\displaystyle
%     \log \frac{x_1}{x_N}, \cdots , \log \frac{x_{N-1}}{x_N}
%   \end{bmatrix}
% \]
% The inverse additive log ratio transform maps values in
% $\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
% \mathbb{R}^{N-1}$ by
% \[
%   \textrm{alr}^{-1}(y)
%   = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
% \]
% where for $u \in \mathbb{R}^N$,

@@

\subsubsection{Isometric Log Ratio Transform}

We may define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform. By noting that the ALR transformation may be written as

@@

The inverse isometric log ratio transform maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{ilr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} \exp(\mathbf{V^{-1}} \mathbf{x} ) \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,
\[
  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
\]

The affine transformation of the multiplication by the inverse unitary matrix does not change
the calculation of the change of variables adjustment proceeds similarly to the ALR case. In fact, any affine transformation may be completed at this stage which generalizes the ILR to power transformations as in \cite{tsagris2011data}\footnote{
   Let $\alpha$ be the scalar power transformation parameter in $\mathbb{R}$ then the inverse \textit{power} based transformation is given by $(V^{-1} \mathbf{x} \alpha + 1) / \alpha$. See \cite{tsagris2011data} for more information.
}.

Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
we can transform to a density over unconstrained parameters $y \in
\mathbb{R}^{N-1}$ by applying the inverse ILR transform and adjusting
for the change of variables, which yields
\[
  p_Y(y) = p_X(\textrm{ilr}^{-1}(y)) \absdet{J_{s}(y)},
\]
where $J_{s}(y)$ is the Jacobian of the function $s$ evaluated at $y$
and $\absdet{J_s(y)}$ is the absolute value of its determinant.

\subsubsection{Augmented-Softmax Transform}
%TODO:  Needs to be written like other transforms -> transform, inverse, |J|
We define the constraining transformation
$f: \mathbb{R}^N \to \Delta^{N} \times \mathbb{R}_{>0}: y \mapsto
(x_-, r)$, where and $x_{-} \in \Delta_{-}^N$. Here $r = \sum_{i=1}^N \textrm{exp}(y_i) > 0$,
$x_i = \frac{1}{r} \textrm{exp}(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function. If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $N$-vector of ones, then
\[
  J_f = (I_{N-1} - x_- \boldsymbol{1}_{N-1}^\top) \operatorname{diag}(x_-),
\]

@@

\subsubsection{Hyperspherical coordinate transforms}

Let $u$ be a unit vector with $N+1$ elements in the positive orthant of the unit sphere $\mathbb{S}_{>0}^N$, which has the constraints $\sum_{i=1}^{N+1} u_i^2 = 1$ and $u_i > 0$.
There is a unique bijective map $f_1\colon \mathbb{S}_{>0}^N \to \Delta^N \colon u \mapsto x$ given by $x_i = u_i^2$.
We then define $f_2$ as the map from hyperspherical coordinates to Cartesian coordinates of the hypersphere.
Taking $z_i = \sin^2(\phi_i) \in (0, 1]$ for $\phi_i \in (0, \frac{\pi}{2}]$, the composed map $f_1 \circ f_2$ is written

\[
  x_i = (1 - z_i (1 - \delta_{i,N+1})) \prod_{k=1}^{i-1} z_k.
\]

The inverse of this transform is
$$z_i = 1 - \frac{x_i}{1 - \sum_{k=1}^{i - 1} x_k}.$$

To complete the transform $f = f_1 \circ f_2 \circ f_3$, we must select a constraining transform $f_3: y \mapsto \phi$.
We consider 3 different approaches.

\subsubsection{Hyperspherical angular transform}

@@

@@

$$
|J_f| = 2^N \prod_{i=1}^N \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin^{2(N-i)+1}(\phi_i) \cos(\phi_i).
$$
This transform was discussed in \cite{betancourt2012cruising}.

\subsubsection{Hyperspherical logit transform}

If we define $f_3$ to map from the unconstrained space directly to a conveniently-chosen power of $z_i$ function using the logistic function 
$$z_i^{N-i+1} = \operatorname{logit}^{-1} (y_i),$$
then the Jacobian determinant is
$$
|J_f| = \frac{1}{N!} \prod_{i=1}^N \operatorname{logit}^{-1}(y_i) (1 - \operatorname{logit}^{-1}(y_i)).
$$
If $x$ is uniformly distributed on the unit simplex, then $y_i$ is logistic-distributed with mean 0 and variance $\frac{\pi^2}{3}$.

\subsubsection{Hyperspherical probit transform}

Similarly, if we use the inverse probit function instead of the inverse logit, then $f_3$ is
$$z_i^{N-i+1} = \operatorname{probit}^{-1} (y_i),$$
where $\operatorname{probit}^{-1}(\cdot)$ is the cumulative distribution function of the standard normal distribution.
The resulting Jacobian determinant is
\[
|J_f| = \frac{1}{N!} \prod_{i=1}^N \mathcal{N}(y_i),
\]
where $\mathcal{N}(\cdot)$ is the density function of the standard normal distribution.
As a result, if $x$ is uniformly distributed on the unit simplex, then $y_i$ is standard normally distributed.

%TODO: better captions
\section{Results}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=1.2\textwidth]{figures/simplex/rmse.png}
    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps. At every iteration in a run, cumulative mean of samples is used to calculate RMSE, which is plotted against the cumulative leapfrog steps till that iteration}
    \label{fig:rmse}
\end{figure}

\subsubsection*{Acknowledgements}

>>>>

In this paper, we focus on transforms on a unit $N$-simplex. Simplexes are useful for representations of multinomial probabilities - points on the standard $N$-simplex in $R^{N+1}$ are the space of possible parameters (probabilities) of the categorical distribution on $n+1$ possible outcomes. Lastly, we define Aitchison Geometry and compare metric distortions of canonical transforms (or their augmented forms) in Compositional Data Analysis.
\section{Theory of constrained parametrizations}

A surjective constraining transformation $f : \mathcal{Y} \rightarrow \mathcal{X}$ is mapped from an unconstrained space $\mathcal{Y} = \mathbb{R}^M$ onto a constrained space $\mathcal{X} \subset \mathbb{R}^N$, and $h(y)$ is a proper density defined over $y \in \mathcal{Y}$ such that if $y \sim h(\cdot)$, then $f(y) \sim \textrm{uniform}(\cdot \mid \mathcal{X})$.  It is not possible to sample over an improper density.  To this end, we stipulate the induced distribution on $y$ to be proper, given a proper density $p_X(x)$ defined for $x \in \mathcal{X}$ such that if $y \sim p_Y(\cdot)$ then $f(y) \sim p_X(\cdot)$ . In this case, density of the distribution induced on $y$ is 
\[
p_Y(y) = h(y) \, p_X(f(y))
\]

@@

 Suppose, we have a distribution $p_X(x)$ on constrained values $x \in \mathcal{X}$. A smooth, bijective unconstraining transformationis defined as $g : \mathcal{X} \rightarrow \mathcal{Y}$, and $g^{-1}$ is the constraining transformation $f : \mathcal{Y} \rightarrow \mathcal{X}$.  Using the standard change of variables, the induced density on $\mathcal{Y}$ is adjusted by a constant factor

@@

For a constraining transformation $f: R^\mathcal{M} \rightarrow R^\mathcal{N}$, if $\mathcal{M} >  \mathcal{N}$, additional constraints are needed to ensure the correct distributions on $\mathcal{X}$. Bijective change-of-variables method cannot be applied here, instead an expanded parameter space $\mathcal{Z} = \mathcal{X} \times \mathcal{R}$ is constructed with parameters $z = (x, r)$. Here, $r \in \mathcal{R}$ is introduced such that $\mathcal{Y}$ and $\mathcal{Z}$ have the same dimensionality. Taking prior $p_R(r | x)$,
\[
p_Y(y) = h(y) p_Z(z) = h(y) p_X(x) p_R(r | x).
\]
We define a bijective transformation $\tilde{g}: \mathcal{Z} \to \mathcal{Y}$ and its inverse $\tilde{f}$, where $\tilde{f}(y)_1 = f(y)$, this permits obtaining change-of-variables,
\[
h(y) = |J_{\tilde{f}}(y)|.
\]
Discarding $r$ after sampling is equivalent to marginalizing it out, and in practice, $r$ may never be directly computed. On account of convenience,  we choose $p_R(r | x) = p_R(r)$ in the rest of the paper.

% In all of these cases, we will have a $p_Y(y)$ defined as a product of the change-of-variables adjustment for the transformationcoupled with a standard normal distribution over $\mathcal{Y}$,
% \[
%   h(y) = \textrm{normal}(y \mid 0, \textrm{I}) \, \absdet{J_f(y)},
% \]  
% where $\textrm{I}$ is the identity matrix.
% For each case, we have to ensure that $p_Y(y) = h(y) \, p_X(f(y))$ is such that if $y \sim p_Y(\cdot)$, then $f(y) \sim p_X(\cdot)$.

\section{Simplex}

@@

Compositional data analysis deals with vectors of strictly positive quantities that adhere to a constant sum constraint. Aitchison\cite{aitchison1982statistical} defined compositional data on a simplical space:
$$
\Delta^N=\left\{\mathbf{x}=\left[x_1, x_2, \ldots, x_D\right] \in \mathbb{R}^N \mid x_i>0, i=1,2, \ldots, N ; \sum_{i=1}^N x_i=\kappa\right\}
$$

@@

\section{Reparametrizations for sampling from a simplex}

\subsection{Additive Log Ratio}
\paragraph{Unconstraining transformation}
The unconstraining transformation for the identified softmax is known as
the additive log ratio (ALR) parametrization
\cite{aitchison1982statistical}, which is a bijection
$\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
$x \in \Delta^{N-1}$ by
\[
  \textrm{alr}(x)
  = \begin{bmatrix}\displaystyle
    \log \frac{x_1}{x_N}, \cdots , \log \frac{x_{N-1}}{x_N}
  \end{bmatrix}
\]
\paragraph{Constraining transformation}
The inverse additive log ratio transformation maps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{alr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,

@@

\paragraph{Change-of-variables adjustment}
\[
\textrm{Here, } \abs{J} = \prod \exp(y)
  \, \left( \frac{1}{1 + \sum(\exp(y))} \right)^N
\]

\subsection{Stick-Breaking}

The stick-breaking transformation is derived from the standard construction for Dirichlet processes \cite{sethuraman1994constructive}. The inverse mapping from unconstrained to constrained parameters involves taking a unit-length stick and recursively breaking pieces off until it is divided into $N + 1$ pieces, the lengths of which provide the entries in the simplex.  Following the transforms used by Stan \cite{stan2022ref}, we add an offset so that the uniform simplex $x = \begin{bmatrix} 1/N & \cdots & 1/N \end{bmatrix}$ maps to $y = \begin{bmatrix} 0 & \cdots & 0 \end{bmatrix}$, a vector of zeros.

\paragraph{Unconstraining transform}
The unconstraining transformation $f : \Delta^{N} \to  \mathbb{R}^N$ is defined so that if $f(x) = y$, then
\[
y_n
= \mathrm{logit}(z_n)
  - \mbox{log}\left(\frac{1}{N + 1 - n} \right),
\]
where 
\[ 
 z_n = \frac{x_i}{1 - \sum_{i = 1}^{n-1} x_{i}}.
\]

\paragraph{Constraining inverse transform}
The inverse transformation follows the stick breaking construction, with $f^{-1} \colon \mathbb{R}^N \to \Delta^N$ defined by $f^{-1}(y) = x$, where
\[
x_n = z_n \left( 1 - \sum_{i=1}^{n-1} x_i \right), \\
\]
with
\[
z_n = \mathrm{logit}^{-1}\!\left(
y_n  + \log \left( \frac{1}{N - n} \right)
\right).
\]

\paragraph{Change-of-variables adjustment} 
The absolute determinant of the Jacobian of the inverse transformationis
\[
\abs{J_{f^{-1}}} 
= \prod_{n=1}^{N-1}
   z_n \, (1 - z_n)
   \left( 1 - \sum_{i=1}^{n-1} x_{i} \right).
\]

% \subsection{Additive log ratio transform}
% The unconstraining transformationfor the identified softmax is known as
% the additive log ratio (ALR) transform
% \cite{aitchison1982statistical}, which is a bijection
% $\textrm{alr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ defined for
% $x \in \Delta^{N-1}$ by
% \[
%   \textrm{alr}(x)
%   = \begin{bmatrix}\displaystyle
%     \log \frac{x_1}{x_N}, \cdots , \log \frac{x_{N-1}}{x_N}
%   \end{bmatrix}
% \]
% The inverse additive log ratio transformationmaps values in
% $\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
% \mathbb{R}^{N-1}$ by
% \[
%   \textrm{alr}^{-1}(y)
%   = \textrm{softmax}(\begin{bmatrix} y &  0 \end{bmatrix}),
% \]
% where for $u \in \mathbb{R}^N$,

@@

\subsection{Isometric Log Ratio Transform}

\paragraph{Unconstraining transformation}
We define a distance preserving transformation from $\textrm{ilr}:\Delta^{N-1} \rightarrow \mathbb{R}^{N-1}$ called an isometric log ratio transform. By noting that the ALR transformation may be written as

@@

\paragraph{Constraining transformation}
The inverse isometric log ratio transformationmaps values in
$\mathbb{R}^{N-1}$ to $\Delta^{N-1}$ defined for $y \in
\mathbb{R}^{N-1}$ by
\[
  \textrm{ilr}^{-1}(y)
  = \textrm{softmax}(\begin{bmatrix} \exp(\mathbf{V^{-1}} \mathbf{x} ) \end{bmatrix}),
\]
where for $u \in \mathbb{R}^N$,
\[
  \textrm{softmax}(u) = \frac{\exp(u)}{\textrm{sum}(\exp(u))}.
\]

The affine transformation of the multiplication by the inverse unitary matrix does not change
the calculation of the change of variables adjustment proceeds similarly to the ALR case. In fact, any affine transformation may be completed at this stage which generalizes the ILR to power transformations as in \cite{tsagris2011data}\footnote{
   Let $\alpha$ be the scalar power transformation parameter in $\mathbb{R}$ then the inverse \textit{power} based transformation is given by $(V^{-1} \mathbf{x} \alpha + 1) / \alpha$. See \cite{tsagris2011data} for more information.
}. 
\paragraph{Change-of-variables adjustment}
Given a density $p_X(x)$ defined over simplices $x \in \Delta^{N-1}$,
we can transform it to a density over unconstrained parameters $y \in
\mathbb{R}^{N-1}$ by applying the inverse ILR transformation and adjusting
for the change of variables, which yields
\[
  p_Y(y) = p_X(\textrm{ilr}^{-1}(y)) \absdet{J_{s}(y)},
\]

\subsection{Augmented-Softmax Transform}
%TODO:  Needs to be written like other transforms -> transform, inverse, |J|
We define the constraining transformation
$f: \mathbb{R}^N \to \Delta^{N} \times \mathbb{R}_{>0}: y \mapsto
(x_-, r)$, where and $x_{-} \in \Delta_{-}^N$. Here $r = \sum_{i=1}^N \textrm{exp}(y_i) > 0$,
$x_i = \frac{1}{r} \textrm{exp}(y_i)$ for $i \in [1, N-1]$ and $\delta_{ij}$ is the Kronecker delta function. If $\mathrm{diag}(x)_{i, j} = \delta_{ij} x_i$ and $\boldsymbol{1}_n$ is the $N$-vector of ones, then
\[
  J_f = (I_{N-1} - x_- \boldsymbol{1}_{N-1}^\top) \operatorname{diag}(x_-),
\]

@@

\subsection{Hyperspherical coordinate transforms}
\paragraph{Unconstraining transformation}
Let $u$ be a unit vector with $N+1$ elements in the positive orthant of the unit sphere $\mathbb{S}_{>0}^N$, which has the constraints $\sum_{i=1}^{N+1} u_i^2 = 1$ and $u_i > 0$.
There is a unique bijective map $f_1\colon \mathbb{S}_{>0}^N \to \Delta^N \colon u \mapsto x$ given by $x_i = u_i^2$.
We then define $f_2$ as the map from hyperspherical coordinates to Cartesian coordinates of the hypersphere.
Taking $z_i = \sin^2(\phi_i) \in (0, 1]$ for $\phi_i \in (0, \frac{\pi}{2}]$, the composed map $f_1 \circ f_2$ is written

\[
  x_i = (1 - z_i (1 - \delta_{i,N+1})) \prod_{k=1}^{i-1} z_k.
\]
\paragraph{Constraining transformation}

The inverse of this transformation is
$$z_i = 1 - \frac{x_i}{1 - \sum_{k=1}^{i - 1} x_k}.$$

To complete the transformation $f = f_1 \circ f_2 \circ f_3$, we must select a constraining transformation $f_3: y \mapsto \phi$.
We consider 3 different approaches.

\subsection{Hyperspherical angular transform}

@@

\paragraph{Change-of-variables adjustment}

@@

$$
|J_f| = 2^N \prod_{i=1}^N \phi_i \left(1 - \frac{2}{\pi} \phi_i\right) \sin^{2(N-i)+1}(\phi_i) \cos(\phi_i).
$$
This transformation was discussed in \cite{betancourt2012cruising}.

\subsection{Hyperspherical logit transform}

We define $f_3$ to map from the unconstrained space directly to a conveniently-chosen power of $z_i$ function using the logistic function 
$$z_i^{N-i+1} = \operatorname{logit}^{-1} (y_i),$$
\paragraph{Change-of-variables adjustment}
$$
|J_f| = \frac{1}{N!} \prod_{i=1}^N \operatorname{logit}^{-1}(y_i) (1 - \operatorname{logit}^{-1}(y_i)).
$$
If $x$ is uniformly distributed on the unit simplex, then $y_i$ is logistic-distributed with mean 0 and variance $\frac{\pi^2}{3}$.

\subsection{Hyperspherical probit transform}

Similarly, if we use the inverse probit function instead of the inverse logit, then $f_3$ is
$$z_i^{N-i+1} = \operatorname{probit}^{-1} (y_i),$$
where $\operatorname{probit}^{-1}(\cdot)$ is the cumulative distribution function of the standard normal distribution.
\paragraph{Change-of-variables adjustment}

\[
|J_f| = \frac{1}{N!} \prod_{i=1}^N \mathcal{N}(y_i),
\]
where $\mathcal{N}(\cdot)$ is the density function of the standard normal distribution.
As a result, if $x$ is uniformly distributed on the unit simplex, then $y_i$ is standard normally distributed.

\section{Metric Geometry of reparametrizations}

Geometrically, coordinates obtained from an Additive Log Ratio transformationdo not correspond to an orthogonal basis, rather it is an oblique basis with respect to the Aitchison metric. Also, the transformationis not symmetrical in all the N components, rather only the first N-1. This distorts euclidean distances measured from the ALR coordinates.
%TODO: better captions
\section{Evaluations}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=1.2\textwidth]{figures/simplex/rmse.png}
    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps. At every iteration in a run, cumulative mean of samples is used to calculate RMSE, which is plotted against the cumulative leapfrog steps till that iteration}
    \label{fig:rmse}
\end{figure}
\subsection{Squared erorr}
\subsection{Effective Sample Size}
\subsection{Transient bias}
\subsection{Autocorrelation time at stationarity}
\subsection{Approximation quality of Laplace at mode in unconstrained  
space}

\section{Conclusion}
\subsection*{Acknowledgements}
